Numerical stability of Euclide algorithm\\over ultrametric fields

### 2016-12-03 resultants/resultants.tex
Small changes

%\documentclass{sig-alternate}
%\documentclass{lms}
\documentclass{article}
\usepackage[utf8]{inputenc}

\usepackage{vmargin}

\usepackage{amsmath,amssymb,amsthm}
%\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{multirow}
\renewcommand*{\theHsection}{\thesection}

@@

\def\todo#1{\ \!\!{\color{red} #1}}
\definecolor{purple}{rgb}{0.6,0,0.6}

@@

\title{Numerical stability of Euclide algorithm\\over ultrametric fields}

%\numberofauthors{1}
%\author{
%\alignauthor Xavier Caruso\\
%  \affaddr{Universit\'e Rennes 1}\\
%  \email{\normalsize \textsf{xavier.caruso@normalesup.org}}
%}
\author{Xavier Caruso}

@@

\begin{abstract}
We address the problem of the stability of the computations of resultants 
and subresultants of polynomials defined over complete discrete valuation 
rings (\emph{e.g.} $\Zp$ or $k[[t]]$ where $k$ is a field). We prove that 
Euclide-like algorithms are highly unstable on average and we explain, in
many cases, how one can stabilize them without sacrifying the complexity.
On the way, we completely determine the distribution of the valuation of
the subresultants of two random monic $p$-adic polynomials having the
same degree.
\end{abstract}

@@

\begin{figure}
\begin{center}
\begin{tabular}{rcl}
$\ring$ &--- & a commutative ring (without any further assumption) \\
$\A$ &---& a complete discrete valuation ring \\
$\pi$ &---& a uniformizer of $\A$ \\
$K$ &---& the fraction field of $\A$ \\
$k$ &---& the residue field of $\A$ \smallskip \\

$\ring_{<n}[X]$ &---& the free $\ring$-module consisting of
polynomials over $\ring$ of degree $<n$ \\
$\ring_{\leq n}[X]$ &---& the free $\ring$-module consisting of
polynomials over $\ring$ of degree $\leq n$ \\
$\ring_n[X]$ &---& the affine space consisting of \emph{monic}
polynomials over $\ring$ of degree $n$. \smallskip \\

$\Res^{d_A,d_B}(A,B)$ &---& The resultant of $A$ and
$B$ ``computed in degree $(d_A,d_B)$'' \\
$\Res_j^{d_A,d_B}(A,B)$ &---& The $j$-th subresultant of $A$ and
$B$ ``computed in degree $(d_A,d_B)$'' \\
\end{tabular}
\end{center}

@@

\begin{ex}
\label{ex:Euclide}
As an illustration, let us examine the behaviour of the precision
on the sequence $(R_i)$ while executing Algorithm~\ref{algo:Euclide} 
with the input:
\begin{align*}
A & = X^5 + 
      \big(27 + O(2^5)\big) X^4 +
      \big(11 + O(2^5)\big) X^3 +
      \big(5 + O(2^5)\big) X^2 +
      \big(18 + O(2^5)\big) X +
      \big(25 + O(2^5)\big) \\
B & = X^5 + 
      \big(24 + O(2^5)\big) X^4 +
      \big(25 + O(2^5)\big) X^3 +
      \big(12 + O(2^5)\big) X^2 +
      \big(3 + O(2^5)\big) X +
      \big(10 + O(2^5)\big).
\end{align*}
The remainder in the Euclidean division of $A$ by $B$ is $S_3 = A-B$.
According to Eq.~\eqref{eq:precsub},
we do not loose precision while performing this substraction and the
result we get is:
$$S_3 = 
      \big(3 + O(2^5)\big) X^4 +
      \big(18 + O(2^5)\big) X^3 +
      \big(25 + O(2^5)\big) X^2 +
      \big(15 + O(2^5)\big) X +
      \big(15 + O(2^5)\big).$$ 
In order to compute $S_4$, we have now to perform the Euclidean division 
of $S_2 = B$ by $S_3$. Noting that the leading coefficient of $S_2$ has 
valuation $0$ and using Eq~\eqref{eq:precadd}--\eqref{eq:precdiv}, we 
deduce that this operation does not loose precision again. We get:
$$S_4 = 
      \big(26 + O(2^5)\big) X^3 +
      \big(17 + O(2^5)\big) X^2 +
      \big(4 + O(2^5)\big) X +
      \big(16 + O(2^5)\big).$$
We observe now that the leading coefficient of $S_4$ has valuation $1$. 
According to Eq.~\eqref{eq:precdiv}, divising by this coefficient ---
and therefore \emph{a fortioti} computing the euclidean division of 
$S_3$ by $S_4$ --- will result in loosing at least one digit in
relative precision. The result we find is:
$$S_5 = 
      \underbrace{\big(\textstyle \frac 3 4 + O(2^2)\big)}_{\text{rel. prec.} = 4} X^2 +
      \underbrace{\big(6 + O(2^3)\big)}_{\text{rel. prec.} = 2} X +
      \underbrace{\big(3 + O(2^3)\big)}_{\text{rel. prec.} = 3}.$$
Continuing this process, we obtain:
$$S_6 = 
      \big(20 + O(2^5)\big) X +
      \big(12 + O(2^5)\big) 
\quad \text{and} \quad
S_7 =
      \textstyle \frac 7 4 + O(2).$$
The relative precision on the final result $S_7$ is then $3$, which is
less than the initial precision which was $5$.
\end{ex}

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-remainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
\begin{equation}
\label{eq:subressequence}
\left\{\begin{array}{ll}
S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\varepsilon_i + 1}
s_{i-1}^{-1} \: c_i^{-\varepsilon_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 0 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\varepsilon_{i+1}} \cdot c_i^{1-\varepsilon_{i+1}}
& \text{for } i \geq -1. 
\end{array}\right.
\end{equation}
Here $n_i = \deg S_i$, $\varepsilon_i = n_{i+1} - n_i$ and $s_i$ is the 
leading coefficient of $S_i$ if $i \geq 0$ and $s_{-1} = 1$ by 
convention. These sequences are finite and the above recurrence applies 
until $S_i$ has reached the value $0$.

@@

\begin{ex}
\label{ex:subres}
Applying Algorithm~\ref{algo:subres} with the input $(A,B)$ of 
Example~\ref{ex:Euclide}, we obtain:
\begin{align*}
R_4 & = 
  \big(29 + O(2^5)\big) X^4 + 
  \big(14 + O(2^5)\big) X^3 + 
  \big(5 + O(2^5)\big) X^2 + 
  \big(17 + O(2^5)\big) X + 
  \big(17 + O(2^5)\big) \\
R_3 & = 
  \big(4 + O(2^5)\big) X^3 + 
  \big(13 + O(2^5)\big) X^2 + 
  \big(4 + O(2^5)\big) X + 
  \big(16 + O(2^5)\big) \\
R_2 & = 
  \big(5 + O(2^5)\big) X^2 + 
  \big(20 + O(2^5)\big) X + 
  O(2^5) \\
R_1 & = 
  \big(1 + O(2)\big) X + 
  \big(1 + O(2)\big) \\
R_0 & = 1 + O(2)
\end{align*}
We observe in particular that the absolute precision on $R_0$ is $1$, 
although it should be at least $5$ since $R_0$ is given by an integral
polynomial expression in terms of the coefficients of $A$ and $B$. We
note moreover that the relative precision on $R_0$ (which is $1$ as
well) is worse that the relative precision we got on $S_7$ (which was
$3$) while executing Algorithm~\ref{algo:Euclide} (\emph{cf} 
Example~\ref{ex:Euclide}).
\end{ex}

@@

\begin{proof}
By Theorem \ref{th:lawVj}, we have $\E[V_j] =
\sum_{i=0}^d \E[Z_i]$ with $Z_i = \min(X_{j-i}, \ldots, X_{j+i})$ ($j$ 
is fixed during all the proof). Our conventions imply that $Z_i$ 
vanishes if $i \geq d-j$. Let us define 
$\tau(1), \ldots, \tau(d-j)$ as the numbers $\sigma(1), \ldots, 
\sigma(d-j)$ sorted in increasing order. For $i < d-j$, the random variable $Z_i$ 
is the minimum of $\tau(i)$ independant random variables with 
geometric distribution of parameter $(1 - q^{-1})$. Thus its 
distribution is geometric of parameter $(1 - q^{-\tau(i)})$.
Its expected value is then $\frac 1 {q^{\tau(i) - 1}}$ and the first
formula follows. The inequality $\frac 1 {q-1} \leq \E[V_j]$ is clear
because $\frac 1 {q-1}$ is the first summand in the expansion of
$\E[V_j]$. The upper bound is derived as follows:
$$\E[V_j] < \sum_{i=0}^\infty \frac 1{q^i - 1}
\leq \sum_{i=0}^\infty \frac 1{q^i - q^{i-1}} = \frac q {(q-1)^2}.$$

@@

By functoriality of subresultants, we may assume that $\ring = \Z[a_0, 
\ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$ are the two 
generic monic polynomials $A = X^d + \sum_{i=0} a_i X^i$ and $B = X^d + 
\sum_{i=0} b_i X^i$. Under this additional assumption, all principal
subresultant are nonzero. Therefore, the sequences $(R_j)_j$, $(U_j)_j$
and $(V_j)_j$ are given by the recurrences \eqref{eq:recRj}--\eqref{eq:recVj}.
The two first announced relations follow easily. Let now focus on the
third one. We set $\tilde R_j = R_j$ and $\tilde R_k = r_j^{2(j-k-1)} 
\: R_k$ for $k < j$. An easy decreasing induction on $k$ shows that this 
sequence obeys to the recurrence:
$$\tilde R_{k-1} = \tilde r_k^2 \cdot \tilde r_{k+1}^{-2} \cdot 
(\tilde R_{k+1} \,\%\, \tilde R_k)$$
where $\tilde r_j = 1$ and $\tilde r_k$ is the coefficient of $\tilde 
R_k$ of degree $k$ for all $k < j$. Comparing with \eqref{eq:recRj},
this implies that $\tilde R_k$ is the $k$-th subresultant of the pair 
$(R_j, R_{j-1})$ and we are done. The fourth equality and the last
statement are proved in a similar fashion.
\end{proof}

@@

To conclude the proof, it remains to prove that the composite in 
the other direction $\psi_j \circ \varphi_j$ is the identity as well. 
Since both $\varphi_j$ and $\psi_j$ are componant-wise given by 
polynomials, we can use functoriality and assume that $\ring$ is 
the field $\Q(c_0, c_1, \ldots, c_n)$ (with $n = 2d$) where
each variable $c_i$ corresponds to one coefficient of $\mathcal U_j$, 
$\mathcal U_{j-1}$, $\mathcal R_j$ and $\mathcal R_{j-1}$ with the
convention that $c_0$ (resp. $(-1)^{j-1} c_0$) is used for the leading
coefficients of $\mathcal R_j$ (resp. $\mathcal U_{j-1}$). Set:
$$\begin{array}{ll}
& (A,B) = \varphi_j(\mathcal U_j, \mathcal U_{j-1}, 
\mathcal R_j, \mathcal R_{j-1}) \smallskip \\
\text{and} & (U_j, U_{j-1}, R_j, R_{j-1}) = \psi_j(A,B)
\end{array}$$
Since $\ring$ is a field and $\mathcal R_j[j]$ does not vanish, the 
Sylvester mapping
$$\begin{array}{rcl}
\ring_{<d-j}[X] \times \ring_{<d-j}[X] & \to & 
\ring_{<2d-j}[X]/\ring_{<j}[X] \smallskip \\
(U,V) & \mapsto & AU + BV
\end{array}$$
has to be bijective. Therefore there must exist $\lambda \in \ring$ 
such that $\mathcal R_j = \lambda \cdot R_j$ and $\mathcal U_j = \lambda 
\cdot U_j$. Similarly $(\mathcal R_{j-1}, \mathcal U_{j-1}) = \mu \cdot
(R_{j-1}, U_{j-1})$ for some $\mu \in \ring$. Identifying the leadings 
coefficients, we get $\lambda = \mu$. Now observe that:
$$\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) = 
\Res^{d-j,d-j-1}(\lambda U_{j-1}, \lambda U_j) = 
\lambda^{2d-2j-1} \cdot \Res^{d-j,d-j-1}(U_{j-1}, U_j).$$
Noting that $(\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, \mathcal 
R_{j-1})$ and $(U_j, U_{j-1}, R_j, R_{j-1})$ both belong to $\mathcal
E_j$, we derive:
$$\mathcal R_j[j]^{2(d-j-1)} = \lambda^{2d-2j-1} \cdot R_j[j]^{2(d-j-1)}$$
from what we finally get $\lambda = 1$ since $\mathcal R_j = \lambda
\cdot R_j$.
\end{proof}

@@

\paragraph{Tracking precision \emph{locally}}

Assume first that we want to perform a given rather simple operation --- 
corresponding, say, to an elementary step (\emph{e.g.} an iteration of 
the main loop) of the algorithm we are executing --- modeled by a 
function $g$ of class $C^1$ defined on an open subset $U$ of a finite 
dimensional normed $K$-vector space $E$ and taking values in another 
finite dimensional normed $K$-vector space $F$.
Our input is an approximated element of $U$ which is represented 
by a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
of the form $C = x+H$ for some $x \in U$. We would like to insist on the 
following: the value of $x$ is a priori \emph{not} given; only is given 
the subset $C$. However, since $H$ is stable under addtion, we have $C = 
x+H$ for \emph{any} element $x \in C$.\footnote{This assertion means that 
any element of the ``rectangle'' $C$ is a center of it... which might be 
surprising if we are accustomed to real numbers.} As explained in \S 
\ref{subsubsec:computDVR}, assuming that $g$ is given as an algebraic 
expression, the naive solution for evaluating $g(C)$ consists in using 
formulae \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
results in an overestimation on the precision, in the following sense: 
this method leads to some inclusion
$$g(C) = g(x+H) \subset y + H_{\text{naive}}$$
where $y \in F$ and $H_{\text{naive}}$ is a lattice which is generally 
much more larger that $g'(x)(H)$, the latter being the best possible
one according to
Lemma~\ref{lem:padicprec} (assuming that the assumptions of this Lemma
are fullfiled). In order to avoid this and be sharp on
precision, another solution consists in splitting the computation of 
$g(C)$ into two parts as follows:
\begin{enumerate}[(A)]
\item \label{item:gpx} compute $g'(x)(H)$, and
\item \label{item:gx} compute $g(x)$ for some $x \in C$.
\end{enumerate}
Part~\eqref{item:gpx} is not easy to handle in full generality: in order 
to be efficient, a special close analysis taking advantage of the 
particular problem under consideration is often necessary. For now, let 
us simply assume that we are given two lattices $H_\min$ and $H_\max$ 
with the property that:
\begin{equation}
\label{eq:Hminmax}
H_\min \subset g'(x)(H) \subset H_\max.
\end{equation}
We shall see later (\emph{cf} \S \ref{subsec:stabilization}) how these
lattices can be constructed --- for a negligible cost --- in the special 
case of subresultants.

@@

\paragraph{Tracking precision \emph{globally}}

As already said, we shall use the above method for tracking precision 
while executing a single step in a complete algorithm. Let us now 
address the problem of ``glueing''. We consider an algorithm \texttt{F} 
consisting in a succession of $n$ steps $\texttt{G}_0, \ldots, 
\texttt{G}_{n-1}$. It is modeled by a function $f : U \to F$ of class 
$C^1$ where $U$ is an open subset in a finite dimensional normed 
$K$-vector space $E$ and $F$ is a finite dimensional normed $K$-vector 
space. The input of \texttt{F} is an approximated element in $U$ 
represented as a coset $C = x+H$ where $x \in U$ and $H$ is a lattice.
We also introduce notations for each individual step. For all $i$, we
assume that $\texttt{G}_i$ is modeled by a function $g_i : U_i \to
U_{i+1}$ of class $C^1$ where $U_i$ is an open subset is 
some normed $K$-vector space $E_i$ and, by convention, $U_0 = U$, $E_0 
= E$ and $U_n = E_n = F$. We thus have:

@@

According to Corollary \ref{cor:Vj}, the expected value of the variable 
$\max(V_0, \ldots, V_{d-1})$ is in $O(\log_p d)$. Thus, the average 
complexity of Algorithm 1 is $O(d^2 \cdot \M(N + \log d))$ bit 
operations. In all usual cases (\emph{cf} Remark \ref{rem:M}), this
complexity is also $\tilde O(d^2 N \cdot \log |k|)$ bit operations.

>>>>

\documentclass{jT}
%\documentclass{article}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb,amsthm}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[norelsize,algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{multirow}
\renewcommand*{\theHsection}{\thesection}

\numberwithin{equation}{section}

@@

\definecolor{purple}{rgb}{0.6,0,0.6}

@@

\title[Numerical stability of Euclide algorithm over ultrametric fields]
{Numerical stability of Euclide algorithm\\over ultrametric fields}

\author{Xavier Caruso}
\address{Xavier Caruso\\
Université Rennes 1, IRMAR, 35042 Rennes Cedex\\
France}
\email{xavier.caruso@normalesup.org}
\urladdr{http://perso.univ-rennes1.fr/xavier.caruso}

\subjclass{11S99, 11Y99, 68W30}

\keywords{Euclide algorithm, ultrametric precision, subresultants}

@@

\begin{resume}
Nous étudions le problème de la stabilité du calcul des résultants et
sous-résultants des polynômes définis sur des anneaux de valuation
discrète complets (\emph{e.g.} $\Zp$ ou $k[[t]]$ où $k$ est un corps).
Nous démontrons que les algorithmes de type Euclide sont très instables
en moyenne et, dans de nombreux cas, nous expliquons comment les rendre
stables sans dégrader la complexité. Chemin faisant, nous déterminons
la loi de la valuation des sous-résultants de deux polynômes $p$-adiques
aléatoires unitaires de même degré.
\end{resume}

\begin{abstr}
We address the problem of the stability of the computations of resultants 
and subresultants of polynomials defined over complete discrete valuation 
rings (\emph{e.g.} $\Zp$ or $k[[t]]$ where $k$ is a field). We prove that 
Euclide-like algorithms are highly unstable on average and we explain, in
many cases, how one can stabilize them without sacrifying the complexity.
On the way, we completely determine the distribution of the valuation of
the subresultants of two random monic $p$-adic polynomials having the
same degree.
\end{abstr}

@@

\begin{figure}
\begin{center}
\begin{tabular}{rcl}
$\ring$ &--- & a commutative ring \\
$\A$ &---& a complete discrete valuation ring \\
$\pi$ &---& a uniformizer of $\A$ \\
$K$ &---& the fraction field of $\A$ \\
$k$ &---& the residue field of $\A$ \smallskip \\

$\ring_{<n}[X]$ &---& the free $\ring$-module consisting of
polynomials\\ & & over $\ring$ of degree $<n$ \\
$\ring_{\leq n}[X]$ &---& the free $\ring$-module consisting of
polynomials\\ & & over $\ring$ of degree $\leq n$ \\
$\ring_n[X]$ &---& the affine space consisting of \emph{monic}
polynomials\\ & & over $\ring$ of degree $n$. \smallskip \\

$\Res^{d_A,d_B}(A,B)$ &---& The resultant of $A$ and
$B$\\ & & ``computed in degree $(d_A,d_B)$'' \\
$\Res_j^{d_A,d_B}(A,B)$ &---& The $j$-th subresultant of $A$ and
$B$\\ & & ``computed in degree $(d_A,d_B)$'' \\
\end{tabular}
\end{center}

@@

\begin{ex}
\label{ex:Euclide}
As an illustration, let us examine the behaviour of the precision
on the sequence $(R_i)$ while executing Algorithm~\ref{algo:Euclide} 
with the input:
\begin{align*}
A & = X^5 + 
      \big(27 + O(2^5)\big) X^4 +
      \big(11 + O(2^5)\big) X^3 \\
& \hspace{1cm} + 
      \big(5 + O(2^5)\big) X^2 +
      \big(18 + O(2^5)\big) X +
      \big(25 + O(2^5)\big) \\
B & = X^5 + 
      \big(24 + O(2^5)\big) X^4 +
      \big(25 + O(2^5)\big) X^3 \\
& \hspace{1cm} + 
      \big(12 + O(2^5)\big) X^2 +
      \big(3 + O(2^5)\big) X +
      \big(10 + O(2^5)\big).
\end{align*}
The remainder in the Euclidean division of $A$ by $B$ is $S_3 = A-B$.
According to Eq.~\eqref{eq:precsub},
we do not loose precision while performing this substraction and the
result we get is:
\begin{align*}
S_3 & = 
      \big(3 + O(2^5)\big) X^4 +
      \big(18 + O(2^5)\big) X^3 \\
& \hspace{1cm} +
      \big(25 + O(2^5)\big) X^2 +
      \big(15 + O(2^5)\big) X +
      \big(15 + O(2^5)\big).
\end{align*}
In order to compute $S_4$, we have now to perform the Euclidean division 
of $S_2 = B$ by $S_3$. Noting that the leading coefficient of $S_2$ has 
valuation $0$ and using Eq~\eqref{eq:precadd}--\eqref{eq:precdiv}, we 
deduce that this operation does not loose precision again. We get:
$$S_4 = 
      \big(26 + O(2^5)\big) X^3 +
      \big(17 + O(2^5)\big) X^2 +
      \big(4 + O(2^5)\big) X +
      \big(16 + O(2^5)\big).$$
We observe now that the leading coefficient of $S_4$ has valuation $1$. 
According to Eq.~\eqref{eq:precdiv}, divising by this coefficient ---
and therefore \emph{a fortioti} computing the euclidean division of 
$S_3$ by $S_4$ --- will result in loosing at least one digit in
relative precision. The result we find is:
$$S_5 = 
      \underbrace{\big(\textstyle \frac 3 4 + O(2^2)\big)}_{\text{rel. prec.} = 4} X^2 +
      \underbrace{\big(6 + O(2^3)\big)}_{\text{rel. prec.} = 2} X +
      \underbrace{\big(3 + O(2^3)\big)}_{\text{rel. prec.} = 3}.$$
Continuing this process, we obtain:
$$S_6 = 
      \big(20 + O(2^5)\big) X +
      \big(12 + O(2^5)\big) 
\quad \text{and} \quad
S_7 =
      \textstyle \frac 7 4 + O(2).$$
The relative precision on the final result $S_7$ is then $3$, which is
less than the initial precision which was $5$.
\end{ex}

@@

This reinterpretation is based on the so-called \emph{subresultant 
pseudo-re\-mainder sequence} which is defined as follows. We pick $A$ and 
$B$ as above. Denoting by $(P \,\%\, Q)$ 
the remainder in the Euclidean division of $P$ by $Q$, we define two 
recursive sequences $(S_i)$ and $(c_i)$ as follows:
\begin{equation}
\label{eq:subressequence}
\left\{\begin{array}{ll}
S_{-1} = A, \, S_0 = B, \, c_{-1} = 1 \smallskip \\
\displaystyle S_{i+1} = (-s_i)^{\varepsilon_i + 1}
s_{i-1}^{-1} \: c_i^{-\varepsilon_i} \cdot (S_{i-1} \,\%\, S_i)
& \text{for } i \geq 0 \smallskip \\
\displaystyle c_{i+1} = s_{i+1}^{\varepsilon_{i+1}} \cdot c_i^{1-\varepsilon_{i+1}}
& \text{for } i \geq -1. 
\end{array}\right.
\end{equation}
Here $n_i = \deg S_i$, $\varepsilon_i = n_{i+1} - n_i$ and $s_i$ is the 
leading coefficient of $S_i$ if $i \geq 0$ and $s_{-1} = 1$ by 
convention. These sequences are finite and the above recurrence applies 
until $S_i$ has reached the value $0$.

@@

\begin{ex}
\label{ex:subres}
Applying Algorithm~\ref{algo:subres} with the input $(A,B)$ of 
Example~\ref{ex:Euclide}, we obtain:
\begin{align*}
R_4 & = 
  \big(29 + O(2^5)\big) X^4 + 
  \big(14 + O(2^5)\big) X^3 + 
  \big(5 + O(2^5)\big) X^2 \\
& \hspace{1cm} + 
  \big(17 + O(2^5)\big) X + 
  \big(17 + O(2^5)\big) \\
R_3 & = 
  \big(4 + O(2^5)\big) X^3 + 
  \big(13 + O(2^5)\big) X^2 + 
  \big(4 + O(2^5)\big) X + 
  \big(16 + O(2^5)\big) \\
R_2 & = 
  \big(5 + O(2^5)\big) X^2 + 
  \big(20 + O(2^5)\big) X + 
  O(2^5) \\
R_1 & = 
  \big(1 + O(2)\big) X + 
  \big(1 + O(2)\big) \\
R_0 & = 1 + O(2)
\end{align*}
We observe in particular that the absolute precision on $R_0$ is $1$, 
although it should be at least $5$ since $R_0$ is given by an integral
polynomial expression in terms of the coefficients of $A$ and $B$. We
note moreover that the relative precision on $R_0$ (which is $1$ as
well) is worse that the relative precision we got on $S_7$ (which was
$3$) while executing Algorithm~\ref{algo:Euclide} (\emph{cf} 
Example~\ref{ex:Euclide}).
\end{ex}

@@

\begin{proof}
By Theorem \ref{th:lawVj}, we have $\E[V_j] =
\sum_{i=0}^d \E[Z_i]$ with 
$$Z_i = \min(X_{j-i}, \ldots, X_{j+i})$$ 
($j$ 
is fixed during all the proof). Our conventions imply that $Z_i$ 
vanishes if $i \geq d-j$. Let us define 
$\tau(1), \ldots, \tau(d-j)$ as the numbers $\sigma(1), \ldots, 
\sigma(d-j)$ sorted in increasing order. For $i < d-j$, the random variable $Z_i$ 
is the minimum of $\tau(i)$ independant random variables with 
geometric distribution of parameter $(1 - q^{-1})$. Thus its 
distribution is geometric of parameter $(1 - q^{-\tau(i)})$.
Its expected value is then $\frac 1 {q^{\tau(i) - 1}}$ and the first
formula follows. The inequality $\frac 1 {q-1} \leq \E[V_j]$ is clear
because $\frac 1 {q-1}$ is the first summand in the expansion of
$\E[V_j]$. The upper bound is derived as follows:
$$\E[V_j] < \sum_{i=0}^\infty \frac 1{q^i - 1}
\leq \sum_{i=0}^\infty \frac 1{q^i - q^{i-1}} = \frac q {(q-1)^2}.$$

@@

By functoriality of subresultants, we may assume that $\ring$ is the
ring $\Z[a_0, 
\ldots, a_{d-1}, b_0, \ldots, b_{d-1}]$ and that $A$ and $B$ are the two 
generic monic polynomials $A = X^d + \sum_{i=0} a_i X^i$ and $B = X^d + 
\sum_{i=0} b_i X^i$. Under this additional assumption, all principal
subresultant are nonzero. Therefore, the sequences $(R_j)_j$, $(U_j)_j$
and $(V_j)_j$ are given by the recurrences \eqref{eq:recRj}--\eqref{eq:recVj}.
The two first announced relations follow easily. Let now focus on the
third one. We set $\tilde R_j = R_j$ and $\tilde R_k = r_j^{2(j-k-1)} 
\: R_k$ for $k < j$. An easy decreasing induction on $k$ shows that this 
sequence obeys to the recurrence:
$$\tilde R_{k-1} = \tilde r_k^2 \cdot \tilde r_{k+1}^{-2} \cdot 
(\tilde R_{k+1} \,\%\, \tilde R_k)$$
where $\tilde r_j = 1$ and $\tilde r_k$ is the coefficient of $\tilde 
R_k$ of degree $k$ for all $k < j$. Comparing with \eqref{eq:recRj},
this implies that $\tilde R_k$ is the $k$-th subresultant of the pair 
$(R_j, R_{j-1})$ and we are done. The fourth equality and the last
statement are proved in a similar fashion.
\end{proof}

@@

To conclude the proof, it remains to prove that the composite in 
the other direction $\psi_j \circ \varphi_j$ is the identity as well. 
Since both $\varphi_j$ and $\psi_j$ are componant-wise given by 
polynomials, we can use functoriality and assume that $\ring$ is 
the field $\Q(c_0, c_1, \ldots, c_n)$ (with $n = 2d$) where
each variable $c_i$ corresponds to one coefficient of $\mathcal U_j$, 
$\mathcal U_{j-1}$, $\mathcal R_j$ and $\mathcal R_{j-1}$ with the
convention that $c_0$ (resp. $(-1)^{j-1} c_0$) is used for the leading
coefficients of $\mathcal R_j$ (resp. $\mathcal U_{j-1}$). Set:
$$\begin{array}{ll}
& (A,B) = \varphi_j(\mathcal U_j, \mathcal U_{j-1}, 
\mathcal R_j, \mathcal R_{j-1}) \smallskip \\
\text{and} & (U_j, U_{j-1}, R_j, R_{j-1}) = \psi_j(A,B)
\end{array}$$
Since $\ring$ is a field and $\mathcal R_j[j]$ does not vanish, the 
Sylvester mapping
$$\begin{array}{rcl}
\ring_{<d-j}[X] \times \ring_{<d-j}[X] & \to & 
\ring_{<2d-j}[X]/\ring_{<j}[X] \smallskip \\
(U,V) & \mapsto & AU + BV
\end{array}$$
has to be bijective. Therefore there must exist $\lambda \in \ring$ 
such that $\mathcal R_j = \lambda \cdot R_j$ and $\mathcal U_j = \lambda 
\cdot U_j$. Similarly $(\mathcal R_{j-1}, \mathcal U_{j-1}) = \mu \cdot
(R_{j-1}, U_{j-1})$ for some $\mu \in \ring$. Identifying the leadings 
coefficients, we get $\lambda = \mu$. Now observe that:
\begin{align*}
\Res^{d-j,d-j-1}(\mathcal U_{j-1}, \mathcal U_j) & = 
\Res^{d-j,d-j-1}(\lambda U_{j-1}, \lambda U_j) \\
& = \lambda^{2d-2j-1} \cdot \Res^{d-j,d-j-1}(U_{j-1}, U_j).
\end{align*}
Noting that $(\mathcal U_j, \mathcal U_{j-1}, \mathcal R_j, \mathcal 
R_{j-1})$ and $(U_j, U_{j-1}, R_j, R_{j-1})$ both belong to $\mathcal
E_j$, we derive:
$$\mathcal R_j[j]^{2(d-j-1)} = \lambda^{2d-2j-1} \cdot R_j[j]^{2(d-j-1)}$$
from what we finally get $\lambda = 1$ since $\mathcal R_j = \lambda
\cdot R_j$.
\end{proof}

@@

\bigskip

\noindent
\textit{Tracking precision locally.}
%
Assume first that we want to perform a given rather simple operation --- 
corresponding, say, to an elementary step (\emph{e.g.} an iteration of 
the main loop) of the algorithm we are executing --- modeled by a 
function $g$ of class $C^1$ defined on an open subset $U$ of a finite 
dimensional normed $K$-vector space $E$ and taking values in another 
finite dimensional normed $K$-vector space $F$.
Our input is an approximated element of $U$ which is represented 
by a coset $C$ with respect to some lattice $H$, that is a subset of $U$ 
of the form $C = x+H$ for some $x \in U$. We would like to insist on the 
following: the value of $x$ is a priori \emph{not} given; only is given 
the subset $C$. However, since $H$ is stable under addtion, we have $C = 
x+H$ for \emph{any} element $x \in C$.\footnote{This assertion means that 
any element of the ``rectangle'' $C$ is a center of it... which might be 
surprising if we are accustomed to real numbers.} As explained in \S 
\ref{subsubsec:computDVR}, assuming that $g$ is given as an algebraic 
expression, the naive solution for evaluating $g(C)$ consists in using 
formulae \eqref{eq:precadd}--\eqref{eq:precdiv}. However, this often 
results in an overestimation on the precision, in the following sense: 
this method leads to some inclusion
$$g(C) = g(x+H) \subset y + H_{\text{naive}}$$
where $y \in F$ and $H_{\text{naive}}$ is a lattice which is generally 
much more larger that $g'(x)(H)$, the latter being the best possible
one according to
Lemma~\ref{lem:padicprec} (assuming that the assumptions of this Lemma
are fullfiled). In order to avoid this and be sharp on
precision, another solution consists in splitting the computation of 
$g(C)$ into two parts as follows:
\begin{enumerate}[(A)]
\item \label{item:gpx} compute $g'(x)(H)$, and
\item \label{item:gx} compute $g(x)$ for some $x \in C$.
\end{enumerate}
Part~\eqref{item:gpx} is not easy to handle in full generality: in order 
to be efficient, a special close analysis taking advantage of the 
particular problem under consideration is often necessary. For now, let 
us simply assume that we are given two lattices $H_\min$ and $H_\max$ 
with the property that:
\begin{equation}
\label{eq:Hminmax}
H_\min \subset g'(x)(H) \subset H_\max.
\end{equation}
We shall see later (\emph{cf} \S \ref{subsec:stabilization}) how these
lattices can be constructed --- for a negligible cost --- in the special 
case of subresultants.

@@

\bigskip

\noindent
\textit{Tracking precision globally.}
%
As already said, we shall use the above method for tracking precision 
while executing a single step in a complete algorithm. Let us now 
address the problem of ``glueing''. We consider an algorithm \texttt{F} 
consisting in a succession of $n$ steps $\texttt{G}_0, \ldots, 
\texttt{G}_{n-1}$. It is modeled by a function $f : U \to F$ of class 
$C^1$ where $U$ is an open subset in a finite dimensional normed 
$K$-vector space $E$ and $F$ is a finite dimensional normed $K$-vector 
space. The input of \texttt{F} is an approximated element in $U$ 
represented as a coset $C = x+H$ where $x \in U$ and $H$ is a lattice.
We also introduce notations for each individual step. For all $i$, we
assume that $\texttt{G}_i$ is modeled by a function $g_i : U_i \to
U_{i+1}$ of class $C^1$ where $U_i$ is an open subset is 
some normed $K$-vector space $E_i$ and, by convention, $U_0 = U$, $E_0 
= E$ and $U_n = E_n = F$. We thus have:

@@

According to Corollary \ref{cor:Vj}, the expected value of 
$\max(V_0, \ldots, V_{d-1})$ is in $O(\log_p d)$. Thus, the average 
complexity of Algorithm 1 is $$O(d^2 \cdot \M(N + \log d))$$ bit 
operations. In all usual cases (\emph{cf} Remark \ref{rem:M}), this
complexity is also $\tilde O(d^2 N \cdot \log |k|)$ bit operations.
